Sorting by Reversals

### 2020-07-11 paper.tex
update paper

\def\code#1{\texttt{\frenchspacing#1}}
\def\padding{\vspace{3em}}

@@

\begin{abstract}
    A reversal is a transformation, reversing the order of elements of a permutation. Sorting by Reversals is the problem of finding the shortest sequence of reversals transforming a permutation to the identity. The length of such a sequence is said to be the reversal distance of that permutation. In this paper we introduce fundamental concepts used in the literature on this topic. In particular, the fundamental relationship between the reversal distance and the number of alternating cycles in a cycle decomposition of a suitably-defined bi-colored graph. Then a $3/2$-approximation is discussed in detail. We also reference further related research.
\end{abstract}

@@

In computational biology a fundamental problem is to identify the evolutionary similarity between genome sequences. By far the most common mutation on genomes is the inversion. A genome can be represented by a permutation. An inversion then simply is the reversal of the substring of a permutation \cite{Kececioglu1995}.

We refer by $\langle S_n, \circ \rangle$ to the symmetric group with $S_n = \{(\pi_1\ \dots\ \pi_n) \mid \{\pi_1, \dots, \pi_n\} = [1,n]\}$. Let $\pi \in S_n$ be a permutation and $\pi_i$ denote $\pi(i)$. The identity permutation $id = (1\ \dots\ n)$ is the permutation mapping every input to itself. A \textit{reversal} $\rho(i,j)$ inverses the substring from $\pi_i$ to $\pi_j$ of a permutation. It is represented by the permutation
\begin{align*}
    (1\ \cdots\ i-1\ j\ j-1\ \cdots\ i+1\ i\ j+1\ \cdots\ n).
\end{align*}
Composing $\pi$ with $\rho(i,j)$ yields $(\pi_1\ \cdots\ \pi_{i-1}\ \pi_j\ \pi_{j-1}\ \dots\ \pi_{i+1}\ \pi_i\ \pi_{j+1}\ \cdots\ \pi_n)$ where the elements $\pi_i, \dots, \pi_j$ have been reversed.

@@

\begin{definition}
Given two permutations $\sigma$ and $\tau$, the \textit{reversal distance problem} is the problem of finding the shortest sequence of reversals transforming $\sigma$ into $\tau$. More formally, we want to find reversals $\rho_1, \dots, \rho_d$ such that
\begin{align*}
    \sigma \circ \rho_1 \circ \cdots \circ \rho_d = \tau
\end{align*}
and $d$ is minimal.
The number of necessary reversals $d$ is called the \textit{reversal distance} between $\sigma$ and $\tau$.
\end{definition}

@@

\begin{definition}
\textit{Sorting by Reversals (MIN-SBR)} then is the problem of determining the reversal distance between some permutation $\pi$ and $id$. The reversal distance is denoted by $d(\pi)$. We say that a sequence of reversals transforming $\pi$ into $id$, \textit{sorts} $\pi$.
\end{definition}

@@

In this paper we give an overview of approximation algorithms for MIN-SBR. In section \ref{sec:breakpoints} we first introduce the notion of breakpoints introduced by \citeauthor*{Kececioglu1995} that allow us to formulate a first lower bound for the reversal distance of $\pi$ and lead to a $2$-approximation \cite{Kececioglu1995}. In section \ref{sec:breakpoint_graph} we define a suitable bi-colored graph first introduced by \citeauthor*{Bafna1996} to find a tighter lower bound \cite{Bafna1996}. In this section, we make a fundamental observation relating the reversal distance to the number of alternating cycles in a cycle decomposition of such a graph. Lastly, we consider a $3/2$-approximation described by \citeauthor{Christie1998} \cite{Christie1998} (Section \ref{sec:32_approximation}). A $1.375$-approximation was later obtained by \citeauthor*{Berman2001}, but is not considered here \cite{Berman2001}.

@@

Now we consider the relationship between two consecutive elements of a permutation.

@@

In an effort to make the identity permutation the only permutation without any breakpoints, we extend our definition of permutations (and the symmetric group) by expanding permutations with one fixed initial $0$ and one fixed trailing $n+1$. In the following we only consider such expanded permutations $\pi = (0\ \pi_1\ \dots\ \pi_n\ n+1) \in S_n$ where $\pi_0 = 0$ and $\pi_{n+1} = n+1$. It is easy to see that for an expanded permutation $\pi$, $b(\pi) = 0$ if and only if $\pi = id$.

@@

\begin{corollary}
Using the previous two observations, \citeauthor*{Kececioglu1995} found a first fundamental lower bound to the reversal distance of a permutation $\pi$ \cite{Kececioglu1995}.
\begin{align*}
    d(\pi) \geq \left\lceil \frac{b(\pi)}{2} \right\rceil.
\end{align*}
\end{corollary}

Based on this first lower bound, \citeauthor*{Kececioglu1995} proposed a simple $2$-approximation running in $O(n^2)$ \cite{Kececioglu1995}.

@@

We now try to find a tighter lower bound by introducing the breakpoint graph of a permutation.

@@

The red edges in the breakpoint graph identify the elements of $\pi$ that have to be split apart and the blue edges identify the elements of $\pi$ that have to be brought together (the missing adjacencies) in the process of sorting the permutation $\pi$. Figure \ref{fig:breakpoint_graph} shows the breakpoint graph of the permutation $(0\ 1\ 3\ 4\ 2\ 5)$.

An \textit{alternating cycle decomposition} of $G(\pi)$ is a collection of edge-disjoint alternating cycles, such that every edge of $G(\pi)$ is contained in exactly one cycle of the collection. We observe that for every breakpoint, a vertice in $G(\pi)$ must have a missing adjacency. And conversely, for every missing adjacency a vertex in $G(\pi)$ must have a breakpoint. Therefore, each vertex in our breakpoint graph has an equal number of incident red and blue edges.

@@

The main result we prove in this section, relates the maximum number of alternating cycles of any cycle decomposition of $G(\pi)$ to the reversal distance of $\pi$.

@@

On the other hand, given a $2$-reversal $\rho$, we know that $\rho$ has to act on two red edges of $G(\pi)$. Identify with $k$ and $l$ the vertices incident to the first red edge and with $m$ and $n$ the vertices incident to the second red edge. Now, because $\rho$ is removes both red edges from the breakpoint graph, we know that either $k \sim m$ and $l \sim n$ or $k \sim n$ and $l \sim m$. Thus, forming a $2$-cycle in $G(\pi)$.
\end{proof}

@@

Using this lower bound, \citeauthor*{Bafna1996} already describe a $\frac{7}{4}$-approximation that is not considered in more detail here. \citeauthor*{Christie1998} was able to formulate a lower bound in terms of the number of $2$-cycles in a maximum alternating cycle decomposition \cite{Christie1998}.

@@

In this section we discuss the $3/2$-approximation proposed by \citeauthor*{Christie1998} in detail \cite{Christie1998}. First, we describe how a sorting sequence of reversals can be obtained from an alternating cycle decomposition of $G(\pi)$ (section \ref{sec:reversal_graph}). In section \ref{sec:cycle_decomposition} we then describe how to find such a cycle decomposition. This then allows us to prove the approximation bound (section \ref{sec:approximation_bound}). Lastly, we describe the full algorithm and analyze its complexity respectively (section \ref{sec:algorithm}).

@@

To obtain a sorting sequence of reversals for a permutation $\pi$ given an alternating cycle decomposition $\mathcal{C}$ of its breakpoint graph $G(\pi)$, \citeauthor*{Christie1998} defines another graph.

@@

Let $u$ be a vertex in $R(\mathcal{C})$. Define $l_b(u)$ and $r_b(u)$ to be the positions in $\pi$ of the leftmost and rightmost elements, respectively, that are incident to the blue edge in $G(\pi)$ represented by $u$. Similarly, $l_r(u)$ and $r_r(u)$ are defined to be the positions in $\pi$ of the leftmost and rightmost red edges that are adjacent to the blue edge represented by $u$. Hereby, the position of a red edge is the position of its leftmost incident vertex. Two vertices $u$ and $v$ are connected with an edge if $l_b(u) < l_b(v) < r_b(u) < r_b(v)$ or $l_b(v) < l_b(u) < r_b(v) < r_b(u)$ or $l_r(u) < l_r(v) < r_r(u) < r_r(v)$ or $l_r(v) < l_r(u) < r_r(v) < r_r(u)$.

@@

We observe that the only alternating cycle decomposition of $G(id)$ is $\mathcal{C} = \emptyset$. And therefore, the only reversal graph of the identity permutation $R(\emptyset)$ consists of $n+1$ isolated blue vertices. In the remainder of this chapter it our goal is to describe a way of transforming the reversal graph of a given alternating cycle decomposition into the reversal graph only consisting of isolated blue vertices with as few reversals as possible.

@@

We now prove a couple of lemmas that we use later.

\begin{lemma}
\label{lem:4}
$R(\mathcal{C})$ contains no isolated blue vertices arising from cycles in $\mathcal{C}$.
\end{lemma}

@@

The following is a simple consequence of the above lemma.

@@

@@

Let $A$ be a connected component of $R(\mathcal{C})$. We denote by $A_u$ the subgraph of $R(\mathcal{C}_u)$ that contains all the vertices of $A$. A sequence of reversals that turns the connected component $A$ of $R(\mathcal{C})$ into an isolated blue vertex, is called an \textit{elemination sequence} for $A$.

\begin{lemma}[\citeauthor*{Christie1998}]
\label{lem:7}
If a connected component $A$ of $R(\mathcal{C})$ is oriented and not an isolated blue vertex, it contains a red vertex $u$ such that $A_u$ is still oriented.
\end{lemma}

@@

\begin{proof}
As $A$ is oriented, we can use lemma \ref{lem:7} to repeatedly find a red vertex $u$ in $A$ such that $A_u$ is still oriented and apply $\rho(u)$.
By lemma \ref{lem:6}, we know that every vertex in $R(\mathcal{C})$ arising from an alternating cycle of $G(\pi)$ is contained within the same connected component. Therefore, every alternating cycle represented by vertices constituting $A$ eventually reduces to a $2$-cycle. Moreover, by lemma \ref{lem:1}, $\rho(u)$ is a $2$-reversal iff $u$ arises from an oriented $2$-cycle.

\end{proof}

@@

\begin{definition}
Given a matching $M$, we construct the \textit{ladder graph} $L(M)$. $L(M)$ has a vertex for every $2$-cycle selected in $M$. Vertices form connected components when the represented $2$-cycles share a blue edge in $G(\pi)$. Such connected components of $L(M)$ are called \textit{ladders}. Figure \ref{fig:ladder} shows part of a breakpoint graph with $2$-cycles that form a ladder.

@@

\begin{theorem}
\label{thm:6}
Let $\pi \in S_n$. Then
\begin{align*}
    d(\pi) \leq b(\pi) - \frac{1}{2} c_2(\pi).
\end{align*}
\end{theorem}

@@

By theorem \ref{thm:4}, we can now sort $\pi$ using at least $k+l+u+m$ $2$-reversals and only $u+v$ $0$-reversals. Therefore
\begin{align*}
    d(\pi) &\leq b(\pi) - k - l - u - m + u + v \\
           &= b(\pi) - k - l - m + v.
\end{align*}
All that is left to show is $k+l+m-v \geq \frac{1}{2} c_2(\pi)$. We know:
\begin{enumerate}
    \item $l+m$ is the number of selected $2$-cycles in unoriented connected components of $R(\mathcal{C})$. $k$ is the total number of $2$-cycles in oriented connected components of $R(\mathcal{C})$. As $\left\lceil \frac{y}{2} \right\rceil + z$ is the number of selected $2$-cycles in $\mathcal{C}$, $k+l+m \geq \left\lceil \frac{y}{2} \right\rceil + z$ holds.
    \item As each connected components of $R(\mathcal{C})$ counted by $v$ represents at least one unoriented $2$-cycle, lemma \ref{lem:5} implies that such a connected component must also represent at least one more alternating cycle in $\mathcal{C}$. As the connected component only represents independent $2$-cycles, we know that it must represent at least two independent $2$-cycles. As $z$ is the number of independent $2$-cycles, $v \leq \left\lfloor \frac{z}{2} \right\rfloor$ holds.
    \item As the set of selected $2$-cycles is not necessarily edge-disjoint, $|M| = y+z \geq c_2(\pi)$.
\end{enumerate}
And therefore
\begin{align*}
    k+l+m-v &\geq \left\lceil \frac{y}{2} \right\rceil + z - v                                          & (1) \\
            &\geq \left\lceil \frac{y}{2} \right\rceil + z - \left\lfloor \frac{z}{2} \right\rfloor     & (2) \\
            &= \left\lceil \frac{y}{2} \right\rceil + \left\lceil \frac{z}{2} \right\rceil \\
            &\geq \frac{1}{2} c_2(\pi).                                                                 & (3)
\end{align*}
\end{proof}

@@

The approximation algorithm with a performance ratio of $\frac{3}{2}$, is outlined in the following.

\begin{algorithm}
    \SetKwInOut{Input}{Input}
    \Input{permutation $\pi \in S_n$}
    construct the matching graph $F(\pi)$\;
    find a maximum cardinality matching $M$ for this graph\;
    construct alternating cycle decomposition $\mathcal{C}$ using this matching\;
    construct the reversal graph $R(\mathcal{C})$\;
    find an elimination sequence for $R(\mathcal{C})$\;
    \caption{Approximation algorithm \cite{Christie1998}}
\end{algorithm}

The matching graph can be constructed in $O(n)$ time. The maximum cardinality matching can be obtained from $F(\pi)$ using the algorithm of Micali and Vazirani in $O(|V||E|^{\frac{1}{2}})$ time \cite{Micali1980} where $|V|$ and $|E|$ are the number of vertices and edges respectively. Therefore, it is possible to find such a matching in $O(n^{\frac{3}{2}})$ time. The alternating cycle decomposition can be obtained from the matching in $O(n)$ time. The reversal graph can be constructed in $O(n^2)$ time. The algorithm to find an elimination sequence of $R(\mathcal{C})$ is outlined below and terminates in $O(n^4)$ time. So the overall time-complexity of the algorithm is $O(n^4)$ \cite{Christie1998}. Using results from \citeauthor*{Kaplan1997}, the time-complexity can be improved to $O(n^2)$ \cite{Kaplan1997}.

\begin{algorithm}
    \SetKwInOut{Input}{Input}
    \Input{reversal graph $R(\mathcal{C})$}
    for each unoriented connected component, apply a reversal represented by a blue vertex\;
    \While{not sorted $\pi$}{
        u := first red vertex\;
        found := false\;
        \While{not found}{
            apply $\rho(u)$ to get $\mathcal{C}_u$\;
            generate $R(\mathcal{C}_u)$\;
            count unoriented connected components in $R(\mathcal{C}_u)$\;
            \eIf{there are no unoriented connected components}{
                found := true\;
            }{
                u := next red vertex\;
            }
        }
    }
    \caption{Finding an elimination sequence \cite{Christie1998}}
\end{algorithm}

>>>>

\def\code#1{\texttt{\frenchspacing#1}}
\def\padding{\vspace{2em}}

@@

\begin{abstract}
    A reversal is a transformation, reversing the order of elements of a permutation. Sorting by reversals is the problem of finding the shortest sequence of reversals transforming a permutation to the identity. The length of such a sequence is said to be the reversal distance of that permutation. In this paper we introduce fundamental concepts used in the literature on this topic. In particular, we introduce the fundamental relationship between the reversal distance and the number of alternating cycles in a cycle decomposition of a suitably-defined bi-colored graph. Then, a $3/2$-approximation is discussed in detail. We also reference further related research.
\end{abstract}

@@

In computational biology a fundamental problem is to determine the evolutionary similarity between genome sequences. By far the most common mutation on genomes is the inversion. A genome can be represented by a permutation. An inversion then simply is the reversal of a substring of a permutation \cite{Kececioglu1995}.

We refer by $\langle S_n, \circ \rangle$ to the symmetric group with $S_n = \{(\pi_1\ \dots\ \pi_n) \mid \{\pi_1, \dots, \pi_n\} = [1,n]\}$. Let $\pi \in S_n$ be a permutation and $\pi_i$ denote $\pi(i)$. The identity permutation $id = (1\ \dots\ n)$ is the permutation mapping every input to itself. A \textit{reversal} $\rho(i,j) \in S_n$ inverses the substring from $\pi_i$ to $\pi_j$ of a permutation. It is represented by the permutation
\begin{align*}
    (1\ \cdots\ i-1\ j\ j-1\ \cdots\ i+1\ i\ j+1\ \cdots\ n).
\end{align*}
Composing $\pi$ with $\rho(i,j)$ yields $(\pi_1\ \cdots\ \pi_{i-1}\ \pi_j\ \pi_{j-1}\ \dots\ \pi_{i+1}\ \pi_i\ \pi_{j+1}\ \cdots\ \pi_n)$ where the elements $\pi_i, \dots, \pi_j$ have been reversed.

@@

\begin{definition}
Given two permutations $\sigma, \tau \in S_n$, the \textit{reversal distance problem} is the problem of finding the shortest sequence of reversals transforming $\sigma$ into $\tau$. More formally, we want to find reversals $\rho_1, \dots, \rho_d \in S_n$ such that
\begin{align*}
    \sigma \circ \rho_1 \circ \cdots \circ \rho_d = \tau
\end{align*}
and $d$ is minimal.
The number of necessary reversals $d$ is called the \textit{reversal distance} between $\sigma$ and $\tau$.
\end{definition}

@@

\begin{definition}
\textit{Sorting by reversals (MIN-SBR)} then is the problem of determining the reversal distance between some permutation $\pi \in S_n$ and $id$. The reversal distance is denoted by $d(\pi)$. We say that a sequence of reversals transforming $\pi$ into $id$, \textit{sorts} $\pi$.
\end{definition}

@@

In this paper we give an overview of approximation algorithms for MIN-SBR. We first introduce the notion of breakpoints given by \citeauthor*{Kececioglu1995} allowing us to formulate a first lower bound for the reversal distance of $\pi$ and leading to a $2$-approximation \cite{Kececioglu1995} (section \ref{sec:breakpoints}). In section \ref{sec:breakpoint_graph} we define a suitable bi-colored graph first introduced by \citeauthor*{Bafna1996} to find a tighter lower bound \cite{Bafna1996}. In doing so, we make a fundamental observation relating the reversal distance to the number of alternating cycles in a cycle decomposition of such a graph. Lastly, we consider a $3/2$-approximation described by \citeauthor{Christie1998} \cite{Christie1998} (Section \ref{sec:32_approximation}). A $1.375$-approximation was later obtained by \citeauthor*{Berman2001}, but is not considered here \cite{Berman2001}.

@@

A very fundamental way of approaching the problem of sorting by reversals is to consider the relationship between two consecutive elements of a permutation.

@@

In an effort to make the identity permutation the only permutation without any breakpoints, we extend our definition of permutations (and the symmetric group) by expanding permutations with one initial $0$ and one trailing $n+1$. In the following we only consider such expanded permutations $\pi = (0\ \pi_1\ \dots\ \pi_n\ n+1) \in S_n$ where $\pi_0 = 0$ and $\pi_{n+1} = n+1$. It is easy to see that for an expanded permutation $\pi$, $b(\pi) = 0$ if and only if $\pi = id$.

@@

\begin{corollary}
Using the previous two observations, \citeauthor*{Kececioglu1995} found a first fundamental lower bound to the reversal distance of a permutation $\pi \in S_n$ \cite{Kececioglu1995}.
\begin{align*}
    d(\pi) \geq \left\lceil \frac{b(\pi)}{2} \right\rceil.
\end{align*}
\end{corollary}

Based on this first lower bound, \citeauthor*{Kececioglu1995} proposed a simple $2$-approximation running in $O(n^2)$ time \cite{Kececioglu1995}.

@@

We now try to find a tighter lower bound by introducing the breakpoint graph of a permutation first defined by \citeauthor*{Bafna1996}.

@@

The red edges in the breakpoint graph identify the elements of $\pi$ that have to be split apart and the blue edges identify the elements of $\pi$ that have to be brought together in the process of sorting the permutation $\pi$ (the missing adjacencies). Figure \ref{fig:breakpoint_graph} shows the breakpoint graph of the permutation $(0\ 1\ 3\ 4\ 2\ 5)$.

An \textit{alternating cycle decomposition} of $G(\pi)$ is a set of edge-disjoint alternating cycles, such that every edge of $G(\pi)$ is contained in exactly one cycle of the set. We observe that for every breakpoint, a vertice in $G(\pi)$ must have a missing adjacency. And conversely, for every missing adjacency a vertex in $G(\pi)$ must form a breakpoint. Therefore, each vertex in our breakpoint graph has an equal number of incident red and blue edges.

@@

The main result we prove in this section, relates the maximum number of alternating cycles of any cycle decomposition of $G(\pi)$ to the reversal distance of $\pi$.\padding

@@

On the other hand, given a $2$-reversal $\rho$, we know that $\rho$ has to act on two red edges of $G(\pi)$. Identify with $k$ and $l$ the vertices incident to the first red edge and with $m$ and $n$ the vertices incident to the second red edge. Now, because $\rho$ removes both red edges from the breakpoint graph, we know that either $k \sim m$ and $l \sim n$ or $k \sim n$ and $l \sim m$. Thus, forming a $2$-cycle in $G(\pi)$.
\end{proof}

@@

Using this lower bound, \citeauthor*{Bafna1996} already describe a $\frac{7}{4}$-approximation that is not considered in more detail here. \citeauthor*{Christie1998} formulated a lower bound in terms of the number of $2$-cycles in a maximum alternating cycle decomposition \cite{Christie1998}.

@@

In this section we discuss the $3/2$-approximation proposed by \citeauthor*{Christie1998} in detail \cite{Christie1998}. First, we describe how a sorting sequence of reversals can be obtained from an alternating cycle decomposition of $G(\pi)$ (section \ref{sec:reversal_graph}). In section \ref{sec:cycle_decomposition} we then describe how to find such a cycle decomposition. This then allows us to prove the approximation bound (section \ref{sec:approximation_bound}). Lastly, we describe the full algorithm and analyze its complexity (section \ref{sec:algorithm}).

@@

To obtain a sorting sequence of reversals for a permutation $\pi$ given an alternating cycle decomposition $\mathcal{C}$ of its breakpoint graph $G(\pi)$, \citeauthor*{Christie1998} defines a new graph.

@@

Let $u$ be a vertex in $R(\mathcal{C})$. Define $l_b(u)$ and $r_b(u)$ to be the positions in $\pi$ of the leftmost and rightmost elements, respectively, that are incident to the blue edge in $G(\pi)$ represented by $u$. Similarly, $l_r(u)$ and $r_r(u)$ are defined to be the positions in $\pi$ of the leftmost and rightmost red edges that are adjacent to the blue edge represented by $u$. The position of a red edge is the position of its leftmost incident vertex. Two vertices $u$ and $v$ are connected with an edge if $l_b(u) < l_b(v) < r_b(u) < r_b(v)$ or $l_b(v) < l_b(u) < r_b(v) < r_b(u)$ or $l_r(u) < l_r(v) < r_r(u) < r_r(v)$ or $l_r(v) < l_r(u) < r_r(v) < r_r(u)$.

@@

We observe that the only alternating cycle decomposition of $G(id)$ is $\mathcal{C} = \emptyset$. And therefore, the only reversal graph of the identity permutation $R(\emptyset)$ consists of $n+1$ isolated blue vertices. In the remainder of this section, our goal is to describe a way of transforming the reversal graph of a given alternating cycle decomposition into the reversal graph only consisting of isolated blue vertices with as few reversals as possible. For that purpose, we first examine the effects of single reversals on the reversal graph.\padding

@@

We now consider a couple of lemmas that will help use prove the approximation bound later.

\begin{lemma}
\label{lem:4}
$R(\mathcal{C})$ contains no isolated blue vertices arising from alternating cycles in $\mathcal{C}$.
\end{lemma}

@@

The following lemma is a simple consequence of the above lemma.

@@

\begin{proof}
Omitted \cite{Christie1998}.
\end{proof}\padding

@@

Let $A$ be a connected component of $R(\mathcal{C})$. We denote by $A_u$ the subgraph of $R(\mathcal{C}_u)$ that contains all the vertices of $A$. A sequence of reversals that turns the connected component $A$ of $R(\mathcal{C})$ into an isolated blue vertex, is called an \textit{elemination sequence} of $A$.

\begin{lemma}
\label{lem:7}
If a connected component $A$ of $R(\mathcal{C})$ is oriented and not an isolated blue vertex, it contains a red vertex $u$ such that $A_u$ is still oriented.
\end{lemma}

@@

\begin{proof}
As $A$ is oriented, we can use lemma \ref{lem:7} to repeatedly find a red vertex $u$ in $A$ such that $A_u$ is still oriented and apply $\rho(u)$.
By lemma \ref{lem:6}, we know that every vertex in $R(\mathcal{C})$ arising from an alternating cycle of $G(\pi)$ is contained within the same connected component. Therefore, every alternating cycle represented by vertices in $A$ eventually reduces to a $2$-cycle. Moreover, by lemma \ref{lem:1}, $\rho(u)$ is a $2$-reversal iff $u$ arises from an oriented $2$-cycle.
\end{proof}

@@

\begin{definition}
Given a matching $M$, we construct the \textit{ladder graph} $L(M)$. $L(M)$ has a vertex for every $2$-cycle selected in $M$. Vertices form connected components when the represented $2$-cycles share a blue edge in $G(\pi)$. Such connected components of $L(M)$ are called \textit{ladders}. Figure \ref{fig:ladder} shows a part of a breakpoint graph with $2$-cycles that form a ladder.

@@

\begin{theorem}
\label{thm:6}
Let $\pi \in S_n$. We are able to find a sorting sequence of at most $b(\pi) - \frac{1}{2} c_2(\pi)$ reversals.
\end{theorem}

@@

By theorem \ref{thm:4}, we can now sort $\pi$ using at least $k+l+u+m$ $2$-reversals and only $u+v$ $0$-reversals. Therefore
\begin{align*}
    d(\pi) &\leq b(\pi) - k - l - u - m + u + v \\
           &= b(\pi) - k - l - m + v.
\end{align*}
All that is left to show is $k+l+m-v \geq \frac{1}{2} c_2(\pi)$. We know:
\begin{enumerate}
    \item $l+m$ is the number of selected $2$-cycles in unoriented connected components of $R(\mathcal{C})$. $k$ is the total number of $2$-cycles in oriented connected components of $R(\mathcal{C})$. As $\left\lceil \frac{y}{2} \right\rceil + z$ is the number of selected $2$-cycles in $\mathcal{C}$, $k+l+m \geq \left\lceil \frac{y}{2} \right\rceil + z$ holds.
    \item As each connected components of $R(\mathcal{C})$ counted by $v$ represents at least one unoriented $2$-cycle, lemma \ref{lem:5} implies that such a connected component must represent at least two alternating cycles in $\mathcal{C}$. As the connected component only represents independent $2$-cycles, we know that it must represent at least two independent $2$-cycles. As $z$ is the number of independent $2$-cycles, $v \leq \left\lfloor \frac{z}{2} \right\rfloor$ holds.
    \item We also have that the selected $2$-cycles in $M$ are not necessarily edge-disjoint on their blue edges, yielding $|M| = y+z \geq c_2(\pi)$.
\end{enumerate}
And therefore
\begin{align*}
    k+l+m-v &\geq \left\lceil \frac{y}{2} \right\rceil + z - v                                          & (1) \\
            &\geq \left\lceil \frac{y}{2} \right\rceil + z - \left\lfloor \frac{z}{2} \right\rfloor     & (2) \\
            &= \left\lceil \frac{y}{2} \right\rceil + \left\lceil \frac{z}{2} \right\rceil \\
            &\geq \frac{1}{2} c_2(\pi).                                                                 & (3)
\end{align*}
\end{proof}

@@

The approximation algorithm with a performance ratio of $\frac{3}{2}$, is outlined in figure \ref{alg:approximation}.

\begin{figure}[htb]
    \begin{algorithm}[H]
        \SetKwInOut{Input}{Input}
        \Input{permutation $\pi \in S_n$}
        construct the matching graph $F(\pi)$\;
        find a maximum cardinality matching $M$ for this graph\;
        construct alternating cycle decomposition $\mathcal{C}$ using this matching\;
        construct the reversal graph $R(\mathcal{C})$\;
        find an elimination sequence for $R(\mathcal{C})$\;
        \caption{Approximation algorithm \cite{Christie1998}}
    \end{algorithm}
    \caption{}
    \label{alg:approximation}
\end{figure}

The matching graph can be constructed in $O(n)$ time. The maximum cardinality matching can be obtained from $F(\pi)$ using the algorithm of Micali and Vazirani in $O(|V||E|^{\frac{1}{2}})$ time \cite{Micali1980} where $|V|$ and $|E|$ are the number of vertices and edges respectively. Therefore, it is possible to find such a matching in $O(n^{\frac{3}{2}})$ time. The alternating cycle decomposition can be obtained from the matching in $O(n)$ time. The reversal graph can be constructed in $O(n^2)$ time. The algorithm to find an elimination sequence of $R(\mathcal{C})$ is outlined in figure \ref{alg:elimination_sequence} and runs in $O(n^4)$ time. So the overall time-complexity of the algorithm is $O(n^4)$ \cite{Christie1998}.

Using results from \citeauthor*{Kaplan1997}, the time-complexity of finding an elimination sequence and therefore the time-complexity of the entire algorithm can be improved to $O(n^2)$ \cite{Kaplan1997}.

\begin{figure}[htb]
    \begin{algorithm}[H]
        \SetKwInOut{Input}{Input}
        \Input{reversal graph $R(\mathcal{C})$}
        for each unoriented connected component, apply a reversal represented by a blue vertex\;
        \While{not sorted $\pi$}{
            u := first red vertex\;
            found := false\;
            \While{not found}{
                apply $\rho(u)$ to get $\mathcal{C}_u$\;
                generate $R(\mathcal{C}_u)$\;
                count unoriented connected components in $R(\mathcal{C}_u)$\;
                \eIf{there are no unoriented connected components}{
                    found := true\;
                }{
                    u := next red vertex\;
                }
            }
        }
        \caption{Finding an elimination sequence \cite{Christie1998}}
    \end{algorithm}
    \caption{}
    \label{alg:elimination_sequence}
\end{figure}
